---
title: "Writing Efficient Code"
editor: source
execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(tictoc)
library(microbenchmark)

options(digits = 2)
```

In this module, you'll learn a bit about how to make your code faster in
general.

# Code speed

Fortunately, we stand on the shoulders of giants. Many skilled computer
scientists have put a lot of time and effort into writing R functions that run
as fast as possible. For you, most of the work to speed up code is simply 
finding the right packages to rely on.

However, there are a few baseline principles that can get you a long way. If
your code takes a long time to run, the reason is often one of these:

-   You are doing a very large computation, relative to the power of your
computer.
-   You are repeating a slow process many times.
-   You are trying to do simple things, but on a very large object.

To speed up the code, without deep knowledge of computer algorithms and inner
workings, you can sometimes come up with clever ways to avoid these pitfalls.

First, though: as you start thinking about writing faster code, you'll need a
way to check whether your improvements actually sped up the code.

#### 📖 Required Reading: [The tictoc and microbenchmark packages](https://www.jumpingrivers.com/blog/timing-in-r/)

#### 📖 Required Reading: [The bench package](https://www.tidyverse.org/blog/2018/06/bench-1.0.1/)

#### 📖 Required Reading: [`lobstr::obj_size()`](https://lobstr.r-lib.org/reference/obj_size.html)

# Tip 1: Avoid larger calculations and memory demands

## Save smaller data frames, if you are going to use them many times

Consider the following dataset:

```{r}
#| echo: true
#| message: false
#| label: load-data

library(fivethirtyeight)

glimpse(congress_age)
```

Suppose we want to do some exploratory analysis on these data. Let's use the
`tic()` and `toc()` functions from the **tictoc** packages to check how much
time it takes to run this code:

```{r}
#| results: false
#| message: false
#| fig-show: hide
#| echo: true
#| label: exploratory-analysis

tic()

congress_age %>%
  filter(chamber == "senate") %>%
  summarize(mean(age))

congress_age %>%
  filter(chamber == "senate") %>%
  ggplot() + 
  geom_histogram(mapping = aes(x = age))

congress_age %>%
  filter(chamber == "senate") %>%
  slice_max(n = 5, order_by = age)

```

```{r}
#| echo: true
toc()
```

These are all three reasonable things to do, and they can't be done in the same
pipeline. But wait - we just made R do the process of filtering to only include
Senators three separate times!

```{r}
#| echo: true
#| results: false

tic()
  congress_age %>%
    filter(chamber == "senate")
```

```{r}
#| echo: true
toc()
```

Instead, how about we filter the data once, save it into an object, and call on
that object:

```{r}
#| echo: true
#| fig-show: hide
#| message: false
#| results: false

tic()

 senate <- congress_age %>%
  filter(chamber == "senate")

senate %>%
  summarize(mean(age))

senate %>%
  ggplot() + 
  geom_histogram(mapping = aes(x = age))


senate %>%
  slice_max(n = 5, order_by = age)
```

```{r}
#| echo: true

toc()
```

That is almost **half** the run time! Yes, we are dealing with very small
increments of time (less than a second), but hopefully you could see how this
might scale to larger calculations. 

# Tip 2: Avoid repeating slow steps

## Avoid for-loops

`for()`-loops are deathly slow in R. If you absolutely must iterate over a
process, rely on the `map()` or `apply()` function families. These functions 
offer the same versatility of a `for()`-loop but are (1) more human readable, 
and (2) much faster because they are written in C. 

::: {.callout-note}
# Why choose `map()` instead of `apply()`? 

The `map()` and `apply()` functions will typically be close to the same speed, 
so the largest  argument in favor of the `map()` family of functions are
their consistent syntax, and their consistent object output. 

The `map()` function **always** returns a list and the `map_XXX()` variants
**always** return a vector of a specified data type. Whereas, `apply()` returns
a vector or a list, `lapply()` returns a list, and `sapply()` returns a vector or
a matrix. 
:::

Here is a motivating example demonstrating the difference in run time between 
two functions:

1. a function which loops over every column in a data frame to extract the data
type of that column
2. a function which uses `map_chr()` to apply the `class()` function to
every column of a data frame

```{r}
#| echo: true
#| label: map-apply-functions

loop_func <- function(df){
  
  # initialize the type object
  type <- rep(NA, ncol(df))
  
  for(i in seq_len(ncol(df))) {
    
    # store the class of the ith column of df into the ith entry of type 
    type[i] <- class(df[,i])
  }
  
  return(type)
}

map_func <- function(df){
  
  type <- map_chr(df, class)
  
  return(type)

  }
```

Okay, now that we've made these two functions, let's see how they compare in 
their run time. Below, we've made a simple albeit large dataframe. The dataframe
has five rows and 100,000 columns and the element of every column is a vector
of 1s. 

```{r}
#| label: simple-large-dataframe
#| echo: true

dat <- as.data.frame(
  matrix(1,
         nrow = 5,
         ncol = 100000)
  )

```

The `microbenchmark()` function from the **microbenchmark** package allows us
to compare the run time for multiple functions:

```{r}
#| label: compare-loops
#| cache: true
#| warning: false
#| echo: true

microbenchmark::microbenchmark(loop_func(dat),
                               map_func(dat),
                               times = 3)
```

Looking at the output, we see substantial differences in the run time of these
two functions. On average, the loop function takes 3,271 milliseconds whereas 
the map function takes 13 milliseconds. 

## Use vectorized functions

Better even than `map()` or `apply()` is not to iterate at all! Writing 
vectorized functions is tough, but do-able in many cases. Here's an example:

{{< video https://youtu.be/ShewFCbPHv8?si=sVpKCo0FF28hw40v >}}

::: {.callout-important}
# Check-in: Efficiency

The function below uses `if()`, `else if()` and `else` to check how a 
number should be classified.

```{r}
#| echo: true
#| label: non-vectorized-function

pos_neg_zero <- function(x){
  if(x > 0){
    return("Greater than 0!")
  } 
  else if (x < 0){
    return("Less than 0!")
  } 
  else {
    return("Equal to 0!")
  }
}
```

These functions (`if()`, `else if()`) are not vectorized, meaning they cannot
accept a vector as an input, which is why the code below returns an error:

```{r}
#| error: true
#| label: error-with-vector-for-non-vectorized-function
#| echo: true

x <- seq(from = -4, to = 6)
pos_neg_zero(x)
```

Since this function is not vectorized, we would need to use `map_chr()` to 
apply the `pos_neg_zero()` function to every element of the vector `x`:

```{r}
#| label: iterating-non-vectorized-function
#| echo: true

map_chr(x, .f = pos_neg_zero)
```

1. Rewrite the `pos_neg_zero()` function so it is vectorized. Meaning, the
function should be capable of accepting a vector as an input. *Hint: The trick 
**is not** adding `map_chr()` into the function body!*

:::

# Tip 3: Use faster existing functions

Because R has so many packages, there are often many functions to perform the
same task. Not all functions are created equal!

## `data.table`

For speeding up work with data frames, no package is better than `data.table`!

{{< video https://www.youtube.com/embed/uueVddWwbkk >}}

#### 📖 Required Reading: [data.table cheatsheet](https://pbs.twimg.com/media/Dwoq0ZQXQAE5lJt.jpg)

::: {.callout-important}
# Check-in: Efficiency

2. Here is some `dplyr` code. Re-write it in `data.table` syntax. 

```{r}
#| echo: true
#| eval: false

congress_age %>%
  group_by(state) %>%
  summarize(mean_age = mean(age))

```

3. Using either the `tictoc` or `microbenchmark` packages perform a speed test 
to see which code is faster, and by how much.
:::

# Tip 4: Only improve what needs improving

> “We should forget about small efficiencies, say about 97% of the time:
> premature optimization is the root of all evil. Yet we should not pass up our
> opportunities in that critical 3%. A good programmer will not be lulled into
> complacency by such reasoning, \[they\] will be wise to look carefully at the
> critical code; but only after that code has been identified.” — Donald Knuth.

Speed and computational efficiency are important, but there can be a trade off. 
If you make a small speed improvement, but your code becomes overly complex, 
confusing, or difficult to edit and test it's not worth it!

Also, consider your time and energy as the programmer. Suppose you have a 
function that takes 30-minutes to run. It relies on a subfunction that takes 
30-seconds to run. Should you really spend your efforts making the subfunction
take only 10-seconds?

The art of finding the slow bits of your code where you can make the most
improvement is called **profiling**.

#### 📖 Required Reading: [Profiling](http://adv-r.had.co.nz/Profiling.html#profiling)

#### 📖 Recommended Reading: [Measuring Performance](https://adv-r.hadley.nz/perf-measure.html)

::: {.callout-tip}
# Skimming is sufficient!

For this reading, I strongly encourage you to skim. The code aspects are mostly
more complicated that we expect in this class; but the general overview of the
principles behind speeding up code is helpful.

In your work, you will most likely "profile" manually, by `tictoc`-ing bits of
your code and so forth, not by these fancy methods.
:::

::: {.callout-important}
# Check-in: Efficiency

4. Why is `vapply()` faster than `sapply()`?

a.  It specifies the output type.
b.  It specifies the input type.
c.  It performs fewer safety checks.
d.  It performs fewer iterations.

5. Here are two other ways to compute the square root of a vector.
Which is faster?  fastest? Use microbenchmarking to test your answers!

```{r}
#| echo: true
#| eval: false
x ^ (1 / 2)
exp(log(x) / 2)
```

6. Why is `mean(x)` slower than `sum(x)/length(x)`?

a.  `sum()` is vectorized, while `mean()` is not.
b.  A mean is a smaller number than a sum, so it needs less memory.
c.  `mean()` calculates the mean twice just in case.
d.  `mean()` was written in an old version of R.

:::

## Optional: Super advanced mode

The following are some suggestions of ways you can level-up your code speed. 
These are outside the scope of this class, but you are welcome to explore them!

-   [Dive deep into the R efficiency rabbit hole...](https://csgillespie.github.io/efficientR/)
-   [Parallelize your computations](http://zevross.com/blog/2019/02/12/dramatically-speed-up-your-r-purrr-functions-with-the-furrr-package/)
-   [Make your code run in C++ with Rcpp](http://adv-r.had.co.nz/Rcpp.html#rcpp)
-   Learn to use [recursion](https://techvidvan.com/tutorials/recursion-in-r/)
-   Learn more about [memory allocation](https://adv-r.hadley.nz/names-values.html) and [garbage collecting](http://adv-r.had.co.nz/memory.html) in R.
