---
title: "Practice Activity Week 4"
format: html
embed-resources: true
editor: source
execute: 
  echo: true
  error: true
  eval: false
---

# Accessing the Practice Activity

Download the template Practice Activity Quarto file here: 
<a href="student/pa-4.qmd" download>`pa-4.qmd`</a>

::: {.callout-important}
Be sure to save the file inside your Week 4 folder of your STAT 431 (or 541)
folder! 
:::

```{r}
#| message: false
#| warning: false

library(tidyverse)
```


# Roundabouts

This week, our activity will explore how roundabouts are used around the world!
You might not be as big of a fan of roundabouts as Dr. T, but these data allow 
us to investigate a fun and unique visualization
technique---[a Stankey diagram](https://en.wikipedia.org/wiki/Sankey_diagram).

A Stankey diagram visualizes data **flows**. Specifically, the visual models the
movement from one state to another. We are going to use a Stankey plot to 
visualize how traffic intersections are transformed into various types of 
roundabouts. 

## The `roundabouts` Package

The [`roundabouts` R package](https://github.com/EmilHvitfeldt/roundabouts) 
provides a simple way to interact with the 
[Roundabouts Database](https://roundabouts.kittelson.com/). You likely do not
have the package installed on your computer, so the first step is to install
the package:

```{r}
#| eval: false
#| label: install-roundabouts

pak::pak("emilhvitfeldt/roundabouts")
```

The `roundabouts` dataframe that is included in the package has an `address`
variable which is a bit untidy. Below is some data cleaning code that will 
clean the `address` variable, and create three new variables: `town_city`, 
`county_area`, and `state_region`. 

```{r}
#| label: roundabouts-data-cleaning
#| message: false
#| warning: false

library(roundabouts)

roundabouts_clean <- roundabouts |>
  mutate(
    # Step 1: Remove strange <![CDATA[ tags from the beginning of the address field
    address = stringr::str_remove(address, stringr::fixed("<![CDATA[")), 
    # Step 2: Remove strang ]]> tags from the end of the address field
    address = stringr::str_remove(address, stringr::fixed("]]>")),
    # Step 3: Isolate the country at the end of the string between the ( ) symbols
    country = stringr::str_extract(address, "\\([^)]+\\)$"),
    # Step 4: Remove the brackets surrounding the country name
    country = stringr::str_remove_all(country, "[\\(\\)]"), 
    # Step 5: Remove the country from the address field
    address2 = stringr::str_remove(address, "\\s*\\([^)]+\\)$")) |>
  # Step 6: Separate address into town, county, state by comma
  separate(address2, 
                  into = c("town_city", "county_area", "state_region"), 
                  sep = ",") |> 
  # Step 7: Fix one instance of California instead of CA
  mutate(state_region = str_trim(state_region, side = "both"), 
         state_region = fct_recode(state_region, 
                                   "CA" = "California")
         ) |> 
  # Step 8: Remove companies from states
  filter(! str_detect(state_region, pattern = "Co.")) |> 
  # Step 9: Grab columns we will be using
  select(town_city, county_area, state_region, country, 
         year_completed, type, previous_control_type, control_type)

```

## Creating a Region Variable

Alright, we now have the `country` of each roundabout, but that seems to 
fine grain for our analysis. If we're interested in how roundabouts are used
around the world, it seems like a `region` variable might be a better fit. 

1. Create a new `continent` column in the dataset, which classifies each 
`country` into one of the five major continents---Americas, Europe, Asia, 
Oceania, Africa.

The `countrycode()` function from the `countrycode` package will be helpful
here! Here is an example of how you can use this function to obtain the 
continent of a given country:

```{r}
#| label: continent-demo
#| eval: false

countrycode::countrycode("Brazil",
            origin = "country.name",
            destination = "continent")
```

```{r}
#| label: add-continent-column
#| echo: false

roundabouts_clean <- roundabouts_clean |> 
  mutate(continent = if_else(country == "Kosovo", 
                             "Europe",
                             countrycode::countrycode(country,
                                                      origin = "country.name",
                                                      destination = "continent")
                             )
         )

```

2. You should have gotten a message about **1** matching error. Meaning, there
was one specific country that  couldn't find the name for. **Modify your code
to manually classify the continent of this country.**


## Exploring Roundabouts

Let's explore the data a bit before we jump into visualizing these data. 

3. What countries have the most roundabouts (in these data)? The fewest
roundabouts?

```{r}
#| label: country-total-roundabouts
#| echo: false

roundabouts_clean |> 
  count(country) |> 
  filter(n == max(n) | n == min(n)) |> 
  rename(`Total Roundabouts` = n, 
         `Country` = country) |> 
  gt::gt()
```

4. In these data, what types of roundabouts are most common?

```{r}
#| label: common-types
#| echo: false

roundabouts_clean |> 
  count(type) |> 
  mutate(prop = n / sum(n)) |> 
  arrange(desc(prop)) |> 
  rename(`Type of Roundabout` = type, 
         `Total Roundabouts` = n, 
         `Proportion` = prop) |> 
  gt::gt() |> 
  gt::fmt_percent(columns = Proportion)
```

5. How does the distribution of types differ by continent?

```{r}
#| label: types-by-continent
#| echo: false

```

6. We've seen a lot of traffic calming circles showing up around SLO. What
states in the US have the largest proportion of traffic calming circles?

```{r}
#| label: prop-calming-circle-in-US
#| echo: false

roundabouts_clean |> 
  filter(country == "United States") |> 
  count(state_region, type) |> 
  group_by(state_region) |> 
  mutate(prop = n / sum(n)) |> 
  ungroup() |> 
  filter(type == "Traffic Calming Circle") |> 
  select(`State` = state_region, 
         `Total Roundabouts` = n, 
         `Percent Calming` = prop) |> 
  arrange(desc(`Percent Calming`)) |> 
  gt::gt() |> 
  gt::fmt_percent(columns = `Percent Calming`) |> 
  gt::tab_header(title = "Percentage of Traffic Calming Circles for Each State's Roundabouts")
```

# The `ggalluvial` Package

Now that we have the data ready, let's explore making Stankey plots with the 
[`ggalluvial` R package](https://corybrunson.github.io/ggalluvial/). 

<!-- previous type -> type (of roundabout) -> control type -->

